---
title: "data_processing_notes"
author: "Rujula Adwant"
date: '2022-07-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to dplyr

dplyr is an R package in the tidyverse. *We can load the package using*

``` {r}

install.packages("dplyr")

#load required libraries
library(dplyr)
library(ggplot2)

#preview dataset
View(starwars)
head(starwars)


```

#### Filtering data 
We can filter data using `filter()`

Tips:
* Be sure you spell the column name correctly (and the value name if it's a categorical variable). Remember, R is case-sensitive.
* Be sure to use a double equal sign when comparing observations; `=` is an assignment operator
* You can use >, <, >=, <= to compare numeric or categorical variables (nominal variables are ranked alphabetically while ordinal variables have a built-in rank)

``` {r}

filter(starwars, hair_color == "blond")
filter(starwars, hair_color == "blond" & species == "Human")
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

filteredData <- filter(starwars, species == "Human", homeworld == "Tatooine")

filter(starwars, skin_color == "gold" & height >= 150)


importantPlanets <- c("Tatooine", "Naboo", "Hoth")
filter(starwars, homeworld %in% importantPlanets)
# %in% is IMPORTANT

# ranked data
head(diamonds) #displays six rows
class(diamonds$cut) #gives you the specific type of data
summary(diamonds$cut) #gives you a count of each category or summary statistics if numeric
summary(diamonds$carat)
head(diamonds$cut) # < - displays first 6 values and levels

filter(diamonds, cut > "Good")

# Ordering categorical data
unique(starwars$eye_color)
factor(starwars$eye_color,
       c("red", "orange", "gold", "yellow", "green,yellow", "blue", "black"), 
         ordered = T)
       
filter(starwars, height < 100, species != "Droid")

#Which characters were born in between the years 100 and 200 (inclusive)
filter(starwars, birth_year <= 200, birth_year >= 100)

## Which characters weight over 100kg but are shorter than 185cm
filter(starwars, mass > 100 & height < 185)

### Missing values 
NA > 5 # it doesn't know
NA == NA

?is.na()


mary_age <- NA
paul_age <- NA
mary_age == paul_age

mass <- NA
mass == 10
10 == NA

is.na(mass)

#### to add NA values, you need to explicitly include them with is.na()

##Which characters are missing a hair color?
filter(starwars, is.na(hair_color))


```

### Arranging data
`arrange()` reorders rows. It does not remove any rows.


```{r}

arrange(starwars, birth_year)

arrange(starwars, desc(birth_year))

### Practice!
## arrange characters to find tallest and shortest

arrange(starwars, height)

# alphabetize by name
arrange(starwars, name)

## How could you use arrange to sort all missing values to the start?
arrange(starwars, desc(is.na(starwars))) #continue for other variables


```

### We can select certain columns in the dataset
`select()` allows us to retain certain variables (columns)
doesn't change order but removes columns not named

``` {r}

select(starwars, hair_color)
head(starwars)
select(starwars, hair_color:eye_color) # returns every column between first:last, including first:last

select(starwars, -hair_color)
select(starwars, -(hair_color:eye_color))

select(starwars, contains("color")) # keeps every column w color
select(starwars, ends_with("color")) # keeps every column that ends with


## Practice!!

## Select out the homeworld and species for the starwars dataset. What question might this subset of variables help answer?

select(starwars, homeworld, species)

## Brainstorm as many ways as possible to select hair_color, eye_color, skin_color, and birth_year

select(starwars, hair_color:birth_year)
```

### Adding new columns
`mutate()` adds new columns to the end of your dataset

```{r}

starwars_small <- select(starwars, height, mass, birth_year)
starwars_small
head(starwars_small)

starwars_small <- mutate(starwars_small, height_m = height/100)
starwars_small
mutate(starwars_small, bmi = mass/(height_m^2))

#to keep only the new columns, use transmute()

# using aggregate functions
mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = T))

```